Supplemental data to ``\emph{k}-mer-based GWAS enhances causal variant and candidate gene discovery in soybean''

### 2023-02-25 additional_files/additional_file_1.tex
Integrate suggestions by JA Sibbesen into the supplemental methods

We used the Illumina WGS data of the 389 selected \emph{G. max} accessions as a
first source of data for SV discovery using methods similar to those used by
\cite{lemay2022}.  Briefly, we used four different SV discovery programs or
pipelines:

@@

\item We used AsmVar \citep[version of 2015-04-16, ][]{liu2015} to call SVs
	from \emph{de novo} assemblies generated with SOAPdenovo2 v. 2.04
		\citep{luo2012} and aligned to the reference genome using LAST
		v. 1047 \citep{kielbasa2011}.
\item We ran Manta v. 1.6.0 \citep{chen2016} in 78 randomly selected batches of
	5 samples and 1 batch of 4 samples. We used the candidate SVs
		identified from each of the batches for downstream analyses.
\item We used smoove v. 0.2.4 \citep{pedersen2019} with the sequence of
	commands \texttt{smoove call}, \texttt{smoove merge}, \texttt{smoove
		genotype} and \texttt{smoove paste} to obtain a dataset of
		candidate SVs. For three (3) of the 389 samples, we were unable
		to obtain candidate variants from smoove due to a reproducible
		cryptic error (segmentation fault) when running \texttt{smoove
		call}. We nevertheless decided to use smoove given that this
		represented a very small fraction of the total number of
		samples and that other programs were able to successfully
		discover SVs from those samples.
\item We used SvABA v. 1.1.3 \citep{wala2018} to call SVs from each of the
	samples separately. We classified the SVs called as breakends by SvABA
		as either deletions, duplications or inversions, and converted
		them to explicit sequence using our own R script. Variants
		called as indels by SvABA were used as is for downstream
		analyses.
\end{enumerate}

We filtered the SVs called by all tools to remove those smaller than 50 bp or
larger than 500 kb in size, those located on unanchored scaffolds or organellar
genomes, those classified as unresolved breakends (\texttt{SVTYPE=BND}), and
any variant containing at least one N in either the reference or alternate
allele sequence.  We converted all variants to a sequence-explicit format and
normalized them using bcftools \texttt{norm} v. 1.10.2-105
\citep{li2009-samtools}.

We generated a single VCF file from the variants called using Illumina data by
running SVmerge \citep{wong2010} in two steps. In the first step, we merged
variants among the ones called by a given tool by running SVmerge using a
sliding window of 5 bp to merge together very similar variants called by the
same tool. In the second step, we used the output of the first merging steps
for all four tools and merged similar variants across these four datasets using
a sliding window of 15 bp in SVmerge.  We proceeded this way because using a
sliding window of 15 bp on the outputs of all tools without pre-merging
resulted in a computationally intractable number of pairwise comparisons.

@@

We used SVs discovered among 17 Canadian soybean cultivars sequenced using
Oxford Nano\-pore Technologies (ONT) by \cite{lemay2022} as a second source of
data for SV discovery.  Although that set of cultivars does not overlap the
ones analyzed in this study, the high quality of the SVs that can be discovered
from long-read sequencing data motivated the inclusion of this dataset among
the SV candidates subjected to genotyping.  We merged the SVs detected in the
long-read (ONT) data with the ones discovered in the short-read (Illumina) data
using SVmerge and systematically favored SVs discovered by Illumina whenever
SVs from the two sets were to be merged.  Our rationale for favoring variants
discovered from the Illumina data was that whenever both technologies discover
the same variants, Illumina data is likely to be more precise given its higher
basecalling accuracy.

@@

In addition to using the Illumina data from 389 accessions, we also used
high-quality genome assemblies of 29 \emph{G. max} or \emph{Glycine soja}
accessions and called SVs by comparing their sequence to that of Williams82. We
retrieved 26 high-quality genomes assembled by \cite{liu2020-pangenome} as well
as that of the cultivar Zhonghuang 13 (ZH13) from the Genome Warehouse
repository (\url{https://ngdc.cncb.ac.cn/gwh/}).  We also retrieved the genomes
of \emph{G. soja} accession W05 and \emph{G. max} cultivar Lee from SoyBase
\citep{grant2010}.

We used methods based on \cite{liu2020-pangenome} to call SVs based on those
genome assemblies.  First, we used the nucmer program of the MUMmer suite of
tools v. 4.0.0rc1 \citep{marcais2018} with option \texttt{-c 1000} to align
each of the assemblies to the Williams82 assembly. Next, we used the
delta-filter program of the MUMmer suite to keep only one-to-one alignments
between any genome and the Williams82 reference using the \texttt{-1} option.
We used these filtered alignments to call SVs using a version of the svmu
program \citep{chakraborty2019} that we forked from the original repository
(\url{https://github.com/malemay/svmu}) in order to improve execution time and
memory usage.

We used a set of tools that we assembled in an R package called svmutools
\url{https://github.com/malemay/svmutools} to convert the svmu output to VCF
format. We limited ourselves to processing deletions and insertions because of
the complexity of resolving other types of variants from whole-genome
assemblies.  Copy number variants and inversions called by svmu were therefore
excluded from downstream analyses, although we did take inverted regions into
account when determining the sequence of insertions in these regions to ensure
their proper representation in the output VCF file. We also filtered the
resulting set of variants to remove those smaller than 50 bp or larger than 500
kb and those with any N in either the reference or alternate allele sequence.
Finally, we normalized the VCF files using \texttt{bcftools norm} and merged
them using SVmerge to obtain a single VCF file of variants discovered through
comparison of high-quality genome assemblies.

@@

We used Paragraph v. 2.4a \citep{chen2019} as one of two SV genotyping programs
to generate a set of SV genotypes for the population of 389 samples for use in
GWAS analyses.  The set of SVs used as input for Paragraph was the result of
merging the set of variants discovered from the Illumina and ONT data with
those discovered from the high-quality assemblies and svmu using SVmerge.
Variants in this candidate set were prepared for input to Paragraph by removing
those closer than 150 bp (the maximum read length in our dataset) to any
chromosome end and padding the variant representation using a script written by
\cite{hickey2020} for that purpose.

We used the recommendations by the authors of Paragraph for running their
program at population scale by running Paragraph individually on each sample.
We computed average sequencing depth for each sample from the output of the
\texttt{samtools coverage} command and used 20 times that value for the
\texttt{-M} parameter of Paragraph.  We ran Paragraph using the
\texttt{multigrmpy} command and merged the resulting VCF files using
\texttt{bcftools merge} in order to get a  single genotype file for the whole
population. This VCF was further filtered by setting the genotype calls made
from less than two reads to missing and by removing variants with minor allele
frequency < 0.02 or a proportion of missing data > 0.5. The resulting dataset
of 186,306 SVs was used for GWAS.

\subsection*{Genotyping structural variation using the vg toolkit}
\label{sv-gwas-sv-genotyping-vg}

We also used the vg toolkit v. 1.41.0 \citep{hickey2020} and its Giraffe
aligner \citep{siren2021} for genotyping our set of candidate SVs on the whole
population. The reason for also using vg is that its variation graph model allows
the integration of SNPs and indels and makes for a more robust and integrated
approach to SV genotyping, even though previous benchmarks comparing those two
programs in soybean showed a reduced sensitivity of the vg toolkit in
genotyping insertions \citep{lemay2022}.

We constructed the variation graph using the \texttt{vg autoindex} command with
the \texttt{-\-workflow giraffe} option. We used the Williams82 reference
assembly version 4 as a basis for the graph, to which we added the set of
candidate SVs described above and the set of 7M SNPs and indels called by
Platypus.  We aligned the trimmed reads of the 741 SRA runs in FASTQ format to
the variation graph using the \texttt{vg giraffe} command with default
parameters. Reads that were still paired after adapter- and quality- trimming
were aligned in paired-end mode, whereas the remaining reads were aligned in
single-end mode. The resulting files in GAM format were used to compute read
coverage on the graph using the \texttt{vg pack} command with options
\texttt{-Q5 -s5}. GAM files belonging to the same accession but originating
from different SRA runs were concatenated to be used as input for the
\texttt{vg pack} command such that a single coverage file per accession was
obtained as output of this step. The command \texttt{vg call} was then used to
call genotypes based on read coverage using the parameters
\texttt{-\-genotype-snarls -\-min-length 50} in order to call all variants
$\geq$ 50 bp supported by the graph.

We merged the genotypes called on all samples using \texttt{bcftools merge} to
obtain a single VCF file for the population. Multiallelic records were split
into biallelic records after merging. Genotype calls made with a genotype
quality (GQ) value < 10 were set to missing, and variants were filtered for a
fraction of missing genotypes $\leq$ 0.5 and a minor allele frequency $\geq$
0.02. The resulting VCF file, which was used for GWAS, contained 51,960 SVs.

\subsection*{Comparison of WGS and SoySNP50K genotypes}
\label{annexe-sv-gwas-soysnp50k}
When we first analyzed the results from the GWAS analysis on all 389 samples,
we found discrepancies between the haplotypes at some loci and the expected
phenotype that should correspond to that haplotype for simple traits such as
flower color and seed coat color. Upon further analysis, we found that a
mismatch between the sequencing data and the identity of the accession was the
most probable cause of these discrepancies. In order to filter out all samples
for which this could be the case, we compared the genotypes of SNPs obtained
from the SoySNP50K array \cite{song2013} to those obtained from the Illumina WGS
data of the 385 samples that had a match in the GRIN database. Briefly, we
retrieved SoySNP50K genotype calls from SoyBase \citep{grant2010} and extracted
all SNPs with a MAF $\geq$ 0.1 in our population. We translated the positions
of these SNPs from assembly version 2 to assembly version 4 of Williams by
finding exact matches of 41-bp sequences surrounding the SNP positions. The
32,852 SNPs for which unambiguous positions could be found on genome assembly
version 4 were genotyped from the mapped WGS reads using \texttt{bcftools
mpileup} and \texttt{bcftools call}. We then used \texttt{bcftools gtcheck} to
compute the concordance of genotype calls between WGS samples and their
expected match in the SoySNP50K genotype dataset.  We identified 24 samples
with < 90\% concordance between the two genotype datasets and removed them from
the dataset, in addition to another sample for which concordance was over 90\%
but poor raw sequencing data quality (\%GC content) suggested contamination.
Following these filtering steps, 363 samples remained for GWAS analyses. The
concordance between WGS data and SoySNP50K genotypes is included in
Additional file 2.

@@

% Figures mentioned only in supplemental data
% Flower color
\ldplot{flower_color}{flower color}{\getvar{flower_color_ldkmers}}

>>>>

We used the Illumina whole-genome sequencing (WGS) data from the 389 selected \emph{G. max} accessions
for SV discovery using methods similar to those described by
\cite{lemay2022}.  Briefly, we used four different SV discovery programs or
pipelines:

@@

\item We used AsmVar \citep[version of 2015-04-16, ][]{liu2015} to call SVs
	from \emph{de novo} assemblies generated with SOAPdenovo2 v. 2.04
		\citep{luo2012} and aligned to the reference genome using LAST
		v. 1047 \citep{kielbasa2011}.
\item We ran Manta v. 1.6.0 \citep{chen2016} in randomly selected batches: 77 batches of
	5 samples and 1 batch of 4 samples. We used the candidate SVs
		identified from each of the batches for downstream analyses.
\item We used smoove v. 0.2.4 \citep{pedersen2019} to obtain a dataset of candidate SVs
	using the following sequence of commands:
	\texttt{smoove call}, \texttt{smoove merge}, \texttt{smoove
		genotype} and \texttt{smoove paste}.
		For three (3) of the 389 samples, we were unable
		to obtain candidate variants due to a reproducible
		error (segmentation fault) when running \texttt{smoove
		call}.
	\item We used SvABA v. 1.1.3 \citep{wala2018} to call SVs from each of the
	samples separately. We classified the SVs called as breakends by SvABA
		as either deletions, duplications or inversions, and converted
		them to explicit sequence using a custom R script. Variants
		called as indels by SvABA were used as is for downstream
		analyses.
\end{enumerate}

We filtered the SVs called by each tool to remove those meeting any of the following conditions:

\begin{itemize}
	\item smaller than 50 bp or larger than 500 kb in size
	\item located on unanchored scaffolds or organellar genomes
	\item classified as unresolved breakends (\texttt{SVTYPE=BND})
	\item containing at least one N in either the reference or alternate allele sequence
\end{itemize}

Next, we converted all filtered variants to a sequence-explicit format using
\texttt{bayesTyperTools convertAllele} \citep{sibbesen2018} and normalized them
using bcftools \texttt{norm} v. 1.10.2-105 \citep{li2009-samtools}.

A single VCF file was generated from the variants called using Illumina data by
running SVmerge \citep{wong2010} in two steps. In the first step, we merged
very similar variants among the ones called by the same tool using a
sliding window of 5 bp. Next, we used the output of the first merging steps
for all four tools and merged similar variants across the four datasets using
a sliding window of 15 bp. This approach was used because a
sliding window of 15 bp on the outputs of all tools without pre-merging
resulted in a computationally intractable number of pairwise comparisons.

@@

We used SVs discovered among 17 Canadian soybean cultivars sequenced using
Oxford Nano\-pore Technologies (ONT) by \cite{lemay2022} as an additional
source of candidate SVs. Although this set of cultivars does not overlap the
ones analyzed in this study, it is likely that many of the SVs occurring
in Canadian germplasm also occur in the germplasm used for this study given that modern soybean
cultivars are derived from a rather small set of progenitors. We merged the SVs detected in the
long-read (ONT) data with the ones discovered in the short-read (Illumina) data
using SVmerge and systematically favored SVs discovered by Illumina whenever
SVs from the two sets were to be merged.  Our rationale for favoring variants
discovered from the Illumina data was that whenever both technologies discover
the same variants, the breakpoints of SVs discovered from Illumina data are
likely to be more precise given the higher basecalling accuracy.

@@

In addition to Illumina data from 389 accessions, we also used
high-quality genome assemblies of 29 \emph{G. max} or \emph{Glycine soja}
accessions to call SVs. We
retrieved 26 high-quality genomes assembled by \cite{liu2020-pangenome} as well
as the genome of the cultivar Zhonghuang 13 (ZH13) from the Genome Warehouse
repository (\url{https://ngdc.cncb.ac.cn/gwh/}). Furthermore, we retrieved the genomes
of \emph{G. soja} accession W05 and \emph{G. max} cultivar Lee from SoyBase
\citep{grant2010}.

We used methods based on \cite{liu2020-pangenome} to call SVs from the
genome assemblies.  First, we used the nucmer program of the MUMmer suite of
tools v. 4.0.0rc1 \citep{marcais2018} with option \texttt{-c 1000} to align
each of the assemblies to the Williams82 assembly. Next, we used the
delta-filter program of the MUMmer suite to keep only one-to-one alignments
between any genome and the Williams82 reference using the \texttt{-1} option.
The filtered alignments were used to call SVs with a version of the svmu
program \citep{chakraborty2019} that we forked from the original repository
(\url{https://github.com/malemay/svmu}) in order to improve execution time and
memory usage.

We used a set of tools that we assembled in an R package called svmutools
\url{https://github.com/malemay/svmutools} to convert the svmu output to VCF
format. We limited ourselves to processing deletions and insertions because of
the complexity of resolving other types of variants from whole-genome
assemblies.  Copy number variants and inversions called by svmu were therefore
excluded from downstream analyses. The
resulting set of variants was filtered to remove those smaller than 50 bp or larger than 500
kb and those with any N in either the reference or alternate allele sequence.
Finally, we normalized the VCF files using \texttt{bcftools norm} and merged
them using SVmerge to obtain a single file with variants discovered through
comparison of high-quality genome assemblies.

@@

We used Paragraph v. 2.4a \citep{chen2019}
to generate a set of SV genotypes from the Illumina data of the 389 samples used in
GWAS analyses.  The set of SVs used as input for Paragraph was generated using SVmerge by
merging variants discovered from the Illumina and ONT data with
those discovered from the high-quality assemblies and svmu.
Variants in this candidate set were prepared for input to Paragraph by first removing
those closer than 150 bp (the maximum read length in our dataset) to any
chromosome end and then padding the variant representation using a script written by
\cite{hickey2020}. The latter step was necessary for the variants to be used
as input to Paragraph.

We ran Paragraph individually on each sample as recommended by the authors.
We computed the average sequencing depth for each sample from the output of the
\texttt{samtools coverage} command and used 20 times that value for the
\texttt{-M} parameter of Paragraph.  We ran Paragraph using the
\texttt{multigrmpy} command and merged the resulting VCF files using
\texttt{bcftools merge} in order to get a  single genotype file for the whole
population. Finally, we filtered the genotyped SVs by setting the genotype calls made
from less than two reads to missing and by removing variants with a minor allele
frequency < 0.02 or a proportion of missing data > 0.5. The resulting dataset
of 186,306 called SVs was used for GWAS.

\subsection*{Comparison of WGS and SoySNP50K genotypes}
\label{annexe-sv-gwas-soysnp50k}
When we first analyzed the results from the GWAS analysis on all 389 samples,
we found discrepancies between the haplotypes at some loci and the expected
phenotype that should correspond to that haplotype for simple traits such as
flower color and seed coat color. Upon further analysis, we found that a
mismatch between the sequencing data and the identity of the accession was the
most probable cause of these discrepancies. In order to identify the samples
for which this could be the case, we compared the genotypes of SNPs obtained
from the SoySNP50K array \cite{song2013} to those predicted from the Illumina WGS
data. We did this for the 385 samples that had a match in the GRIN database. Briefly, we
retrieved SoySNP50K genotype calls from SoyBase \citep{grant2010} and extracted
all SNPs with a MAF $\geq$ 0.1 in our population. We then translated the positions
of these SNPs from assembly version 2 to assembly version 4 of Williams by
finding exact matches of 41-bp sequences surrounding the SNP positions. The
32,852 SNPs for which unambiguous positions could be found on genome assembly
version 4 were genotyped from the mapped WGS reads using \texttt{bcftools
mpileup} and \texttt{bcftools call}. We then used \texttt{bcftools gtcheck} to
compute the concordance between the WGS genotype calls and the
SoySNP50K genotype dataset. We identified 24 samples
with < 90\% concordance between the two genotype datasets and removed them from
the dataset. We also excluded an additional sample for which concordance was over 90\%
but poor raw sequencing data quality (\%GC content) suggested contamination.
The concordance between WGS data and SoySNP50K genotypes is included in
Additional file 2 and shown in Figure \ref{concordance-histogram}.

@@

% Figures mentioned only in supplemental data

% Histogram with concordance rates
\begin{cfigure}
	\includegraphics[width = 6in]{concordance_histogram}
	\caption{Histogram of concordance rates between genotypes derived from WGS data and SoySNP50K genotypes for 385 soybean accessions.
	The vertical line indicates the 0.9 concordance threshold that was used for filtering out mismatching samples.}
	\label{concordance-histogram}

\end{cfigure}

% Flower color
\ldplot{flower_color}{flower color}{\getvar{flower_color_ldkmers}}
